From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-26 appendixB/appendixB.Rmd
WIP

which for historical reasons [see @vanKampen2007] is known as a master equation. From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$. These calculations are greatly simplified by observing the following property of the step operator: for any pair of test functions $f_n$, $g_n$, 
\begin{align}
\sum_{n=0}^{N-1}  g_n \mathbb E f_n = \sum_{n=1}^N f_n \mathbb E^{-1} g_n \nonumber
\end{align}

@@

While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.  Moreover, we have yet to make any moment closure argument that solving for the first two moments in this manner will be sufficient.  We will illustrate how this is done in both the traditional diffusion approximation and then in the van Kampen system size expansion.

@@

@Kramers1940 @Moyal1949, @Kurtz1978

@vanKampen1961 @vanKampen1976 @Kurtz1972 @Kurtz1978

A derivation of diffusion approximation in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.  This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.  The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

@@

## The System size expansion

 The method is based on derivation presented by the van Kampen [@vanKampen2007], though a formal proof can be found in earlier work of Thomas Kurtz [@Kurtz1970; @Kurtz1971; @Kurtz1972].

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an *ordinary* differential equation (ODE)
for the macroscopic process, coupled to a *linear* SDE which
governs the *deviations* ($\xi$)  from the macroscopic equation:

@@

### Introduction: a change of variables

@@

We will change Eq \eqref{master1} into the variables $\phi$ and $\xi$.   We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the steop operator under the original discrete variable $n$:

@@

\begin{align}
\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = \alpha_1(\phi) \label{macroscopic}
\end{align}

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen2007, we will call this difference $\alpha_1(\phi)$ as a shorthand\footnote{This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size: $\alpha_{i,j} = \int r^i \Phi_j(r) \ud r$.}

Collecting terms of order $\Omega^0$ we recover the diffusion equation:

\begin{align}
\frac{\partial \Pi}{\partial t} = - \alpha'_{1,0}(\phi) \frac{\partial}{\partial \xi} \xi \Pi + \tfrac{1}{2}\alpha_{2,0}(\phi) \frac{\partial^2}{\partial \xi^2} \Pi \label{FP}
\end{align}

Where we define $\alpha_2(\phi) = b(\phi)+d(\phi)$ as an analogously.  From this it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),

\begin{align}
\partial_t \langle \xi \rangle &= \alpha'_{1,0}(\phi) \langle \xi \rangle \\
\partial_t  \langle \xi^2 \rangle &= 2\alpha'_{1,0}(\phi) \langle \xi \rangle + \alpha_{2,0}(\phi) \label{fluc} 
\end{align}

@@

Eq \eqref{fluc} is particularly instructive.  Note that $\alpha_2(\phi)$ is always positive, and hence will try to increase the fluctuations $\langle \xi^2 \rangle$.  As this term increases, it increases the influence of its coefficient $\alpha_1'(\phi)$.  If this term is negative (as it must be near a stable equilibrium) then it serves to dissipate these fluctuations.  Note the dissipation comes from the macroscopic behavior alone, and does not depend on knowing $b$ and $d$ separately.  Since this dissipation is strongest for large fluctuations and small for very small fluctuations, $\xi^2$ will exponentially approach an equilibrium:

\begin{align}
\langle \xi^2 \rangle = \frac{-\alpha_2(\phi)}{2 \alpha_1'(\phi)} \label{fluctuation dissipation}
\end{align}

However, $\alpha_1'(\phi)$ need not be negative everywhere, but will be negative for any stable point, $\alpha_1'(\phi^s) < 0$.  Hence there will be a region around any stable point where this fluctuation-dissipation relationship given by \eqref{fluctuation dissipation} provides a good description of the fluctuations.  Consequently, for any birth-death process $b>d$ we can write down the equilibrium fluctuations as:

@@

\begin{align}
\alpha_1(\phi) &= c \phi(1- \phi) - e\phi \\
\alpha_2(\phi) &= c \phi(1-\phi) + e\phi
\end{align}

@@

# An alternate derivation: the diffusion approximation

@@

\begin{align*}
\frac{\ud \phi}{\ud t} &= \alpha_{1,0}(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_{1,0}(\psi) 
\end{align*}

@@

\begin{align}
\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial \alpha_{1,0}}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \xi \eta \rangle + \alpha_{2,0} \label{phivar} \\
\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial \alpha_{1,0}}{\partial \phi} + \frac{\partial \beta_{1,0} }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial \alpha_{1,0} }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_{1,0}}{\partial \psi} \langle \eta^2 \rangle + \beta_{2,0} \label{psivar}
\end{align}

Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_{1,0}}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_{2,0}}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.  A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.

@@

\begin{equation*}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} \alpha_{1,0}(\phi,\psi) \right)^2}{\left( \partial_{\phi} \alpha_{1,0}(\phi,\psi) \right)^2 + \partial_{\phi} \alpha_{1,0} (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation*}

From which we recover the environmental variation equation presented in the main text.

@@

Notably, both the SDE derived from the Kramer-Moyal diffusion approximation and the central limit theorem result arising from the system size expansion break down in the case of large deviations.  @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case.

>>>>

which for historical reasons [see @vanKampen2007] is known as a master equation. Note that we can define the master equation more generally in terms of a generic transition probability $w(y|x)$ of going into any state $y$ given that the system is currently in state $x$:

$$
\begin{equation}
\frac{\mathrm{d}}{\mathrm{d}t} P(x,t) = \int w(x|y) P(y,t) - w(y|x)P(x,t) \mathrm{d} x
\end{equation}
$$

The master equation can be thought of as a more process-based description of the Chapman-Kolmorogov equation defining a stochastic process [@vanKampen2007].

<!--
From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$. These calculations are greatly simplified by observing the following property of the step operator: for any pair of test functions $f_n$, $g_n$, 
\begin{align}
\sum_{n=0}^{N-1}  g_n \mathbb E f_n = \sum_{n=1}^N f_n \mathbb E^{-1} g_n \nonumber
\end{align}

@@

While all moments can be derived as above, it is often impossible to solve these equations for nonlinear models.  Moreover, we have yet to make any moment closure argument that solving for the first two moments in this manner will be sufficient.  We will illustrate how this is done in both the traditional diffusion approximation and then in the van Kampen system size expansion.   
-->

@@

A derivation of diffusion approximation is originally credited to @Kramers1940 and later improved by @Moyal1949, though Einstein makes implicit use of this approximation in his own work on Brownian Motion decades earlier [@Gardiner2009].  The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskeinen.  This diffusion approximation is nicely summarized @Gardiner2009 and in @vanKampen2007, though I believe @Kurtz1978 provides a more formal proof that goes unmentioned by any of these references.  Both the diffusion approximation and the van Kampen expansion start with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

@@

## The system size expansion

This method is based on derivation originally proposed by van Kampen @vanKampen1961 @vanKampen1976 and later grounded in more formal work of Thomas Kurtz, [@Kurtz1970; @Kurtz1971; @Kurtz1972; @Kurtz1978].  In contrast to the diffusion approximation above, the system size expansion derives an *ordinary* differential equation (ODE) for the macroscopic process, coupled to a *linear* SDE which governs the *deviations* ($\xi$)  from the macroscopic equation:

@@

 Since this system of equations defines a Gaussian probability distribution, the main text summarizes this result as coupled ODEs for mean and variance, which should be familiar to readers not familiar with stochastic differential equation SDE formulations.  I summarize the derivation here for the birth death process, though an accessible presentation can found for both this and the more general form of the master equation in van Kampen's popular text, @vanKampen2007.

@@

We will change Eq \eqref{master1} into the variables $\phi$ (the macrosopic value, i.e. density) and $\xi$ (i.e. fluctuations/deviations from the expected density).   We begin with the step operator, which can be approximated by a Taylor expansion.To formulate a Taylor expansion of the step operator (where $k$ can be a positive or negative integer), first consider the steop operator under the original discrete variable $n$:

@@

\begin{align}
\frac{\ud \phi(t)}{\ud t} = b(\phi)-d(\phi) = a_1(\phi) \label{macroscopic}
\end{align}

Where we have the part of the birth and death functions that depend on the macroscopic variable $\phi$ alone.  This is known as the macroscopic equation, and corresponds to the density equations commonly written down.  Following @vanKampen2007, we will call this difference $a_1(\phi)$ as a shorthand.  This notation suggests it is the first jump moment, defined as first moment of the transition rate between states in a Markov process.  It so happens this term will give the macroscopic law for any Markov process, not only a birth-death process.  The second moment of the transition rates, $a_2$, will be for us simply the sum of the birth and death rates.  We use this notation as it generalizes to processes that have a different master equation from \eqref{master1}, to describe a transition of arbitrary step size:

$$
\begin{equation}
a_i = \int r^i \Phi(r) \mathrm{d} r
\end{equation}
$$
The first jump moment, $i == 1$ the two possible steps $r$ are $\pm 1$, and thus $a_1$ is sum of the two possible one-step transitions $+1 \cdot b(x) + -1 \cdot d(x)$.  The second jump moment the $r$ term is raised to the second power, making both contributions positive such that $a_2(x) = b(x) + d(x)$

Collecting terms of order $\Omega^0$ we recover the *linear* Fokker Planck equation:

\begin{align}
\frac{\partial \Pi}{\partial t} = - a'_1(\phi) \frac{\partial}{\partial \xi} \xi \Pi + \tfrac{1}{2}a_2(\phi) \frac{\partial^2}{\partial \xi^2} \Pi \label{FP}
\end{align}

Where we define $a_2(\phi) = b(\phi)+d(\phi)$ as an analogously.  Contrast this to the Fokker Planck equation of the diffusion equation, where the birth and death terms appear *inside* the derivatives and the derivatives are with respect to the state variable $x$ rather than the fluctuations, $\xi$.

From here it is a straight forward exercise to calculate the moments of the distribution (multiply by $\xi$, $\xi^2$ and integrate),

\begin{align}
\partial_t \langle \xi \rangle &= a'_1(\phi) \langle \xi \rangle \\
\partial_t  \langle \xi^2 \rangle &= 2a'_1(\phi) \langle \xi \rangle + a_2(\phi) \label{fluc} 
\end{align}

@@

Eq \eqref{fluc} is particularly instructive.  Note that $a_2(\phi)$ is always positive, and hence will try to increase the fluctuations $\langle \xi^2 \rangle$.  As this term increases, it increases the influence of its coefficient $a_1'(\phi)$.  If this term is negative (as it must be near a stable equilibrium) then it serves to dissipate these fluctuations.  Note the dissipation comes from the macroscopic behavior alone, and does not depend on knowing $b$ and $d$ separately.  Since this dissipation is strongest for large fluctuations and small for very small fluctuations, $\xi^2$ will exponentially approach an equilibrium:

\begin{align}
\langle \xi^2 \rangle = \frac{-a_2(\phi)}{2 a_1'(\phi)} \label{fluctuation dissipation}
\end{align}

However, $a_1'(\phi)$ need not be negative everywhere, but will be negative for any stable point, $a_1'(\phi^s) < 0$.  Hence there will be a region around any stable point where this fluctuation-dissipation relationship given by \eqref{fluctuation dissipation} provides a good description of the fluctuations.  Consequently, for any birth-death process $b>d$ we can write down the equilibrium fluctuations as:

@@

\begin{align}
a_1(\phi) &= c \phi(1- \phi) - e\phi \\
a_2(\phi) &= c \phi(1-\phi) + e\phi
\end{align}

@@

@@

\begin{align*}
\frac{\ud \phi}{\ud t} &= a_1(\phi, \psi) \\
\frac{\ud \psi}{\ud t} &= \beta_1(\psi) 
\end{align*}

@@

\begin{align}
\frac{\ud \langle \xi^2 \rangle}{\ud t} &= 2 \frac{\partial a_1}{\partial \phi} \langle \xi^2 \rangle + 2 \frac{\partial a_1 }{\partial \psi} \langle \xi \eta \rangle + a_2 \label{phivar} \\
\frac{\ud \langle \xi \eta \rangle}{\ud t} &= \left( \frac{\partial a_1}{\partial \phi} + \frac{\partial \beta_1 }{\partial \psi} \right)\langle \xi \eta \rangle + \frac{\partial a_1 }{\partial \psi} \langle \eta^2 \rangle \label{covar}\\
\frac{\ud \langle \eta^2 \rangle}{\ud t} &= 2 \frac{\partial \beta_1}{\partial \psi} \langle \eta^2 \rangle + \beta_2 \label{psivar}
\end{align}

Without an explicit model for the environmental variation process, we can merely summarize its role in terms of the autocorrelation, $\tau_c := \tfrac{\partial \beta_1}{\partial \psi}$ and its overall variance at steady-state, $\langle \eta^2 \rangle = \tfrac{\beta_2}{-2 \partial_{\psi} \beta{1,0}}$, which we will refer to as the environmental variation $\sigma^2_e := \langle \eta^2 \rangle$.  A more detailed mechanistic description of how the environmental noise arises would permit a time-dependent solution, but the system size expansion tells us that these two summary statistics are all we need to express the steady-state solution above.

@@

\begin{equation*}
\langle xi^2 \rangle = \frac{ \left( \partial_{\psi} a_1(\phi,\psi) \right)^2}{\left( \partial_{\phi} a_1(\phi,\psi) \right)^2 + \partial_{\phi} a_1 (x,y) \tau_c } \sigma_e^2 + \sigma^2_d
\end{equation*}

<!-- FIXME -->

$$
\begin{align}
\frac{\mathrm{d} x}{\mathrm{d} t} &= \alpha_1(x, y)  \\
\frac{\mathrm{d} y}{\mathrm{d} t} &= \beta_1(y) 
\end{align}
$$

For a generic birth death process that depends both on
on population state $x$ but also on environmental
factors $y$, where we write $a_1(x,y) := b(x,y) - d(x,y)$,
sum $a_2(x,y) := b(x,y) + d(x,y)$

For a generic population model
that depends not only on population state $x$ but also on environmental
factors $y$, it is possible to derive the steady-state variance
(Appendix B) as:

$$
\begin{align}
&\sigma_x^2  = \frac{  \partial_y a_1(x,y) ^2}{ \partial_x a_1(x,y) ^2 +  \tau_c \phantom\cdot \partial_x a_1(x,y)  } \sigma_e^2 + \frac{1}{2} \frac{a_2(x,y)}{\partial_x a_1(x,y)}
\end{align}
$$

where $\sigma_e^2$ is the variance in the environmental noise and
$\tau_c$ the auto-correlation of that noise.

@@

Notably, both the SDE derived from the Kramer-Moyal diffusion approximation and the central limit theorem result arising from the system size expansion break down in the case of large deviations.  @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case, which is of particular interest in the persistence / coexistence literature where stochastic extinction arises only from such large deviations.

### 2018-02-26 paper/paper.Rmd
WIP

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N}$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of varation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma_^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] = 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$.

@@

This result has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological intepretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{\vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskeinen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

- Link back to Levins, show where same
- IMPORTANCE of fluctuation dissipation theorem to the tranisition probability discussion!!
- ENV noise and colored noise

The change in variance can be seen
as a balance between the stochastic contribution, $b(x) + d(x)$
and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
The deterministic contribution is the slope of the macroscopic
equation, that is, the eigenvalue indicating the stability of the
equilibrium point.  For stable points, this value must be negative
and precisely balance the strictly positive stochastic contribution 
of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
the greater the fluctuations).  This balance at steady-state is known as the
fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
expression in understanding stochastic dynamics.  Because Eq \eqref{vanKampen}
also describes the time-evolution of these fluctuations, we can similarly
solve for the correlation between two points separated by time $\tau$.

For our Levins' model, $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$,
we find the average number of occupied patches 
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases only as $\sqrt{N}$ while the steady state itself increases
in direct proportion to $N$, resulting in the smaller relative size
of fluctuations in the larger system in Figure \ref{gillespie}.  We can also see
that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with different levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

Whereas demographic noise reflects processes at a lower scale (the
individual, rather than the population), environmental stochasticity
reflects processes whose dynamics we are not explicitly modelling.
Once again, this can best be illustrated through an explicit
formulation of the master equation.  For a generic population model
that depends not only on population state $x$ but also on environmental
factors $y$, it is possible to derive the steady-state variance
(Appendix B) as:

$$
\begin{equation}
\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d 
\label{env_noise}
\end{equation}
$$

where $\sigma_e^2$ is the variance in the environmental noise and
$\tau_c$ the auto-correlation of that noise, $\sigma_d^2$ the
equilibrium variance contributed by demographic noise from before,
($\tfrac{b + d}{-2\partial_x (b - d)}$) and where we have defined
$f(x,y) = b(x,y) - d(x,y)$ for notational convenience.  For example,
if extinction rate $e$ in our Levins' model is allowed to vary
stochastically as Gaussian white noise with variance $\sigma_e^2$
and mean $e$, we find the expected variance introduced in the
population to be:

@@

Note that the contribution of environmental stochasticity is
independent of system size and directly proportional to the variance
in the environment.  This derivation provides a useful illustration
of the difference between demographic noise and environmental noise:
the latter simply describes the variation introduced by some other
dynamic variable that we are not modelling explicitly, and whose
dynamics are independent of the state of our system, $x$. This
environmental variation no doubt also arises from some more detailed
lower level process, but we need not be explicit about those dynamics
-- the expansion tells us that it is sufficient to know the overall
variance and the degree of auto-correlation.  Recent work has also
highlighted the importance of including both demographic and
environmental noise [@Melbourne2008; @Bonsall2004].  The system
size expansion still requires the assumption that the noise is small
relative to the system size.  In the case of large fluctuations,
this approximation must be replaced with a different approach, such
as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of $\tau_c$, or the degree of auto-correlation, which
emerges from our derivation above, has also been and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work often assumed environmental noise would be
uncorrelated as a mathematical convenience.  As the derivation of
Eq \eqref{env_noise} makes clear, for a continuous-time process,
the relevant metric is the timescale of auto-correlation in the
environmental noise process relative to the auto-correlation timescale
of the population dynamics.  The Gaussian noise that emerges out
of the system-size approximation is always auto-correlated to some
degree -- if we measure our environment at consecutive times
infinitesimally far apart, we expect nearly identical values.  At
intervals much longer than the correlation time, consecutive
measurements are effectively independent (i.e. white noise). 
The degree of auto-correlation in  noise is all about _relative_ timescales.

## Other forms of stochasticity

>>>>

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N}$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of varation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma_^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and nonlinearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

The expected auto-correlation, $\langle x_0, x_t \rangle$ also follows from the
system size expansion, e.g. at stationary state:

$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
which will also prove useful later on.  Note this expression simply states that the equilbrium
variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
the fluctuation-disspation theorem) and that correlations fall off exponentially at a rate given by this slope.   
We will see both autocorrelation and variance play similar roles to infer this eigenvalue when we 
reach the role of noise as informer.

@@

This better-known approach has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological intepretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{\vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskeinen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

Applying the system size approximation to the example of the Levins model introduced
previously lets us both recover the classic model while adding some nuance in 
the interpretation.
Taking $b(n) = n (1 - \tfrac{n}{N})$ and $d(n) = en$ and solving for the steady
state to van Kampen's ODEs, we see the average number $\langle n \rangle$  of occupied patches is
$\langle n \rangle = N\tfrac{e}{c}$ with variance at steady-state of 
$\sigma^2_n = N \tfrac{e}{c}$.  Observe that the variance scales proportionally
to the system size, $N$, which explains the standard deviation
increases as $\sqrt{N}$, as in the canonical model and as predicted by Figure \ref{gillespie}.  
We can also see that different ways of partitioning the same macroscopic equation
into components of birth and death will create the same mean dynamics
but with *different* levels of noise.  For instance, in a logistic
model with $b = rn$ and $d = rn^2/N$, we instead find the variance
at steady state is $\sigma^2 = K$; significantly larger than in the
Levins model.

@@

The van Kampen expansion can also be useful in illustrating how 
environmental noise enters a model as a separate process or dimension.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated rather 
than white noise (see derivation, Appendix B).  This result is intuitive:
for continuous time models, all fluctations are positively auto-correlated
by definition -- measure any continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation makes it clear,
it is in fact the *relative* time scales of the environmental process
and population level process that really matter.  The degree of coupling
to environmental noise also depends on the derivative of the population 
dynamics with respect to the stochastically varying environmental variable:
the larger that derivative, the stronger the contribution of environmental
fluctuations to population fluctuations. For example, if extinction
rate $e$ in our Levins' model is allowed to vary
stochastically as Gaussian white noise with variance $\sigma_e^2$
and mean $e$, we find the expected variance introduced in the
population to be:

@@

Note that the contribution of environmental stochasticity is
independent of system size and directly proportional to the variance
in the environment.

This
environmental variation no doubt also arises from some more detailed
lower level process, but we need not be explicit about those dynamics
-- the expansion tells us that it is sufficient to know the overall
variance and the degree of auto-correlation.  Recent work has also
highlighted the importance of including both demographic and
environmental noise [@Melbourne2008; @Bonsall2004].  The system
size expansion still requires the assumption that the noise is small
relative to the system size.  In the case of large fluctuations,
this approximation must be replaced with a different approach, such
as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of auto-correlation, has and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work often assumed environmental noise would be
uncorrelated as a mathematical convenience.

<!-- This can be condensed or dropped to save space -->

## Origins of other forms of stochasticity
